Fast Sorting on Modern Computers

### 2005-08-30 docs/tech_report/conclusions.tex
so close to being done i can taste it

Among them is the fact that bimodal branch predictors have performed better than
2-level adpative ones in almost every case in this research. This holds across
all comparative branches in all the sorts we surveyed, and we believe that where
explicit patterns - which can be exploited by 2-level adaptive predictors - do
not exist, a bimodal predictors perform better.

@@

\section{Branch Prediction Results}
%TODO mergre this section

It is also shown that bimodal branch predictors generally out-perform two-level
adaptive predictors when resolving comparative misses. This is shown by the
results of insertion sort, bubblesort, shakersort, heapsort and quicksort. In
the cases where the two-level adaptive predictor does perform as well, the table
required is much larger than that required by the bimodal predictors to get the
same results.

@@

%TODO claims that it backed
% TODO claims that it refuted or found problems with
% - cache in heapsort
% - algorithm N, can those improvements really be made

Several discoveries were made and analysed: insertion sort has one misprediction
per key; selection sort has $H_N-1$ branch mispredictions per key - an
explanation of this was also provided; tiling mergesort can be done in two
steps, reducing the cache miss count; it is possible to avoid using a
sentinel and still remove the bounds check from a standard 2-heap, simply by
reordering the steps involved; sequential searches are far cheaper than binary
searches across small lists, due to a low branch misprediction rate.

%TODO add novel contributions
% - avoid copying back in mergesort
% - remove bounds check without sentinel in base heapsort
% - double-tiling

>>>>

Among them is the fact that bimodal branch predictors have performed better than
2-level adaptive ones in almost every case in this research. This holds across
all comparative branches in all the sorts we surveyed, and we believe that where
explicit patterns - which can be exploited by 2-level adaptive predictors - do
not exist, a bimodal predictors perform better.

@@

@@

Several discoveries were made and analysed: insertion sort has one misprediction
per key; selection sort has $H_N-1$ branch mispredictions per key, which applies
to long arrays, though short arrays become more predictable the longer they
get, as in heapsort; tiling mergesort can be done in two steps, reducing the
cache miss count; aligning mergesort for a level 2 cache alone is
counter-productive - aligning it for level 1 and level 2 is significantly
faster; it is possible to avoid using a sentinel and still remove the bounds
check from a standard 2-heap, simply by reordering the steps involved;
sequential searches are far cheaper than binary searches across small lists, due
to a low branch misprediction rate. Bubblesort has terrible branch performance
in the middle, due to being partially sorted - it performs much worse than
selection sort as a result; multi-mergesort contains regular access patterns
which can be exploited by a 2-level adaptive branch predictor; in almost all
other cases, a bimodal predictor out-performs a 2-level adaptive predictor. A
median-of-3 quicksort outperforms other medians; radix sort has no branch
mispredictions; shellsort has easily calculable branch mispredictions, and
performs very similarly to an $O(NlogN)$ so.

In addition, we provide an analysis of most major sorts currently used, a
comparison of $O(NlogN)$ sorts, of different medians for quicksort, of $O(N^2)$
sorts and of cache conscious sorts. We present branch prediction performance
figures for most major sorts, and adapt cache-conscious strategies from
mergesort to make an improved radixsort the fastest performing sort in our
tests.
